Does Anybody Really Know What Time It Is?\\ Automating the Extraction of Date Scalars

### 2015-07-06 conclusion.tex
Word => TeX formatting
Authors, fixed-width fonts, citations, dumb quotes.

While validating the algorithms on a large corpus, we also found that the number of distinct formats in the wild is surprisingly high, and demonstrates the wisdom of including general-purpose date parsing functions in data visualization tools, data cleaning tools and RDBMSes. In particular, it is interesting to note that the most prominent open source RDBMSes (MySQL and Postgres) both have a built-in version of DATEPARSE, possibly reflecting that this is a common need that gets implemented when users are empowered to add to a code base.

>>>>

While validating the algorithms on a large corpus, we also found that the number of distinct formats in the wild is surprisingly high, and demonstrates the wisdom of including general-purpose date parsing functions in data visualization tools, data cleaning tools and RDBMSes. In particular, it is interesting to note that the most prominent open source RDBMSes (MySQL and Postgres) both have a built-in version of \texttt{DATEPARSE}, possibly reflecting that this is a common need that gets implemented when users are empowered to add to a code base.

### 2015-07-06 intro.tex
Word => TeX formatting
Authors, fixed-width fonts, citations, dumb quotes.

In recent years, there has been a growth of interest in data visualization technologies for human-assisted data analysis using systems such as [1,10,11]. While computers can provide high-speed and high-volume data processing, humans have domain knowledge and the ability to process data in parallel, notably by using our visual systems. Most importantly, humans provide the definition of what is valuable in an analysis. Accordingly, human/computer analytic systems are essential to extracting knowledge of value to humans and their societies from the large amounts of data being generated today.

\subsection{Interactivity}
Visualization systems are most effective when they are interactive, thereby allowing a user to explore data and connect it to their domain knowledge and sense of what is important without breaking cognitive flow. In recent years, a number of such systems have been developed, both by the academic community and by the commercial sector. Exploration of data consists not only in creating visual displays, but also in creating and modifying domain-specific computations. Consequently, most data visualization systems include facilities for defining such calculations as part of the data model being analyzed. The most effective systems allow users to define these calculations as part of the analytic interaction, which permits the user to stay in the flow of analysis [9].

During the analytic process, a user may discover that parts of the data are not yet suitable for analysis. Solutions to this problem are often provided by data preparation tools external to the visual analysis environment, which requires the user to break their analytic flow, launch another tool and reprocess their data before returning to their analysis. If the user does not own this process (e.g. it is the responsibility of another department), then there can be significant delays (including ``never.") More subtly, the result of updated external processing may not be compatible with the user's existing work, which can lead to more time lost reconciling the new data model with the existing analysis.

From the user's perspective, the boundary between preparation and analysis is not nearly so clean cut. Bad data is often discovered using visual analysis techniques (e.g. histograms or scatter plots) and it is most natural for the user to ``clean what she sees" instead of switching to a second tool. This leads to an ``adaptive" process whereby users will prefer leveraging existing tools in the analytics environment (no matter how well suited to the task) over switching to another application. Thus a well-designed interactive visual analysis environment will provide tools that enable users to perform such cleaning tasks as interactively as possible.

\subsection{The Tableau Ecosystem}
In this paper, we shall be looking at an example of this problem in the context of the Tableau system. Tableau is a commercial visual analysis environment derived from the Polaris [1] system developed at Stanford. In addition to an interactive desktop application for creating data visualizations, the Tableau ecosystem also includes servers for publishing and sharing interactive visualizations. These servers can be privately operated by individual customers or hosted in the cloud.

\subsubsection{Tableau Public}
Of particular relevance in the present work is a free version of the server environment called Tableau Public [12]. Tableau Public (or ``Public") allows authors to publish data visualizations that can be shared with the general public. The published data sets are limited to 100K rows and must be stored using the Tableau Data Engine (or ``TDE") [2, 3], but are otherwise free to use the full analytic capabilities of the Tableau system. In particular, the TDE provides native support for all analytic calculations defined by the authoring and publishing components. Moreover, visualizations published to Public are uploaded as complete Tableau workbooks, including the data model developed as part of the analysis. This makes it a rich source of data on the analytic habits and frustrations of Tableau users.

@@

\subsection{DATEPARSE}
Among the recent additions to the Tableau function library is a function called DATEPARSE, the usability of which is the focus of the work in this paper.

\subsubsection{	Scalar Dates}
The SQL-99 standard defines three temporal scalar types: DATE, TIMESTAMP and TIME. They are typically implemented as fixed-point types containing an offset from some epoch (\textit{e.g.} Julian Days.) This makes them compact to store and allows some temporal operations to be implemented very efficiently using arithmetic. Thus from the RDBMS perspective, representing dates in scalar form provides numerous benefits for users, both in terms of available operations and query performance.

Tableau models the first two of these types as ``Date" and ``Date \& Time"; the third (pure time) is folded into Date \& Time by appending it to a fixed date of 1899-12-30. From the analytic perspective, date types are dimensional (\textit{i.e.} independent variables) and can be used as either categorical (simply ordered) or quantitative (ordered with a distance metric) fields.
Categorical dates have a natural hierarchy associated with them generated by calendar binning. The visualization in Figure I1 shows an example of a bar chart employing binned categorical dates in a year/quarter hierarchy.

\subsubsection{Parsing}
One of the oldest data preparation problems observed in Tableau is the parsing of date scalars so that users can perform these kinds of analyses. Training materials from the early days of the company included examples of how to convert columns of integers in the form yyyyMMdd to date scalars. The documented solution at the time was to convert the integer to a string and perform some locale-dependent string operations before casting the string back to a date. Unfortunately, this approach had a number of problems:
\begin{itemize}
\item String operations are notoriously slow compared to scalar operations (typically 10-100x slower)
\item Default parsing of date formats is locale-dependent, and may not work when the workbook is shared across an international organization (e.g. between the US and European offices)
\item The expression code was hard to understand and maintain because it used a verbose, general-purpose string-handling syntax instead of a domain language.
\end{itemize}

@@

Our first solution to this problem was to add a new function to the Tableau calculation language called DATEPARSE. This function would take a string column and convert it to a datetime using a special purpose domain language for describing dates. Such functions exist in a number of databases (e.g. Oracle, Postgres and MySQL) and adding it to the TDE was straightforward, so it was a natural addition to the function library.
We chose the date parsing syntax defined by the International Components for Unicode (or ICU) project [8] for the common domain language because it mirrored what was available in the Tableau code base, and translating this syntax into the formats used by various database vendors (e.g. MySQL, Oracle, Postgres) was relatively painless. There are a few patches of non-overlapping functionality, but they tend to be obscure and we provide warnings when functionality is missing.

\subsubsection{	Usability}
After shipping this new function to our user base, we investigated how it was being applied in the field. We examined a few months of workbooks that had been uploaded to Public after the release to see if and how it was being used. We found that although there were a number of new calculations using DATEPARSE, the error rate for the domain language syntax was about 15\%. DATEPARSE was capable of solving the problem, and some users were able to discover it, but the syntax did not appear to be easy to use reliably.

\subsection{Automated Format Extraction}
One solution might have been to design a graphical environment that enabled users to construct valid patterns. This would have involved a substantial development investment with no guarantee that the result would be correct if the user misunderstood the environment. Instead, we have developed two algorithms for automating the derivation of the format string, each of which uses a different machine learning technique. Both algorithms result in over 90\% parsing accuracy – and 100\% syntactic correctness because they are machine generated.

% vidya: this probably should be its own section

\subsection{Related Work}
Data preparation has been a known analytic bottleneck since at least the description of the Potter’s Wheel system [4]. Since then, several other interactive data preparation systems have been proposed [5,6]. While effective, these systems all make assumptions about possible date formats (e.g. the domain system in [4]), which we suggest are too restrictive for real world data.
The Minimum Descriptive Length technique was first proposed in [7]. We have built on the version described in [4].
Related work on parsing languages, outlier detection.

>>>>

In recent years, there has been a growth of interest in data visualization technologies for human-assisted data analysis using systems such as \cite{Polaris,Qlik,Spotfire}. While computers can provide high-speed and high-volume data processing, humans have domain knowledge and the ability to process data in parallel, notably by using our visual systems. Most importantly, humans provide the definition of what is valuable in an analysis. Accordingly, human/computer analytic systems are essential to extracting knowledge of value to humans and their societies from the large amounts of data being generated today.

\subsection{Interactivity}
Visualization systems are most effective when they are interactive, thereby allowing a user to explore data and connect it to their domain knowledge and sense of what is important without breaking cognitive flow. In recent years, a number of such systems have been developed, both by the academic community and by the commercial sector. Exploration of data consists not only in creating visual displays, but also in creating and modifying domain-specific computations. Consequently, most data visualization systems include facilities for defining such calculations as part of the data model being analyzed. The most effective systems allow users to define these calculations as part of the analytic interaction, which permits the user to stay in the flow of analysis \cite{Morton}.

During the analytic process, a user may discover that parts of the data are not yet suitable for analysis. Solutions to this problem are often provided by data preparation tools external to the visual analysis environment, which requires the user to break their analytic flow, launch another tool and reprocess their data before returning to their analysis. If the user does not own this process (e.g. it is the responsibility of another department), then there can be significant delays (including "never.") More subtly, the result of updated external processing may not be compatible with the user's existing work, which can lead to more time lost reconciling the new data model with the existing analysis.

From the user's perspective, the boundary between preparation and analysis is not nearly so clean cut. Bad data is often discovered using visual analysis techniques (e.g. histograms or scatter plots) and it is most natural for the user to "clean what she sees" instead of switching to a second tool. This leads to an "adaptive" process whereby users will prefer leveraging existing tools in the analytics environment (no matter how well suited to the task) over switching to another application. Thus a well-designed interactive visual analysis environment will provide tools that enable users to perform such cleaning tasks as interactively as possible.

\subsection{The Tableau Ecosystem}
In this paper, we shall be looking at an example of this problem in the context of the Tableau system. Tableau is a commercial visual analysis environment derived from the Polaris \cite{Polaris} system developed at Stanford. In addition to an interactive desktop application for creating data visualizations, the Tableau ecosystem also includes servers for publishing and sharing interactive visualizations. These servers can be privately operated by individual customers or hosted in the cloud.

\subsubsection{Tableau Public}
Of particular relevance in the present work is a free version of the server environment called Tableau Public \cite{Public}. Tableau Public (or "Public") allows authors to publish data visualizations that can be shared with the general public. The published data sets are limited to 100K rows and must be stored using the Tableau Data Engine (or "TDE")\cite{TDE1,TDE2}, but are otherwise free to use the full analytic capabilities of the Tableau system. In particular, the TDE provides native support for all analytic calculations defined by the authoring and publishing components. Moreover, visualizations published to Public are uploaded as complete Tableau workbooks, including the data model developed as part of the analysis. This makes it a rich source of data on the analytic habits and frustrations of Tableau users.

@@

\subsection{DATEPARSE}
Among the recent additions to the Tableau function library is a function called \texttt{DATEPARSE}, the usability of which is the focus of the work in this paper.

\subsubsection{	Scalar Dates}
The SQL-99 standard defines three temporal scalar types: \texttt{DATE}, \texttt{TIMESTAMP} and \texttt{TIME}. They are typically implemented as fixed-point types containing an offset from some epoch (\textit{e.g.} Julian Days.) This makes them compact to store and allows some temporal operations to be implemented very efficiently using arithmetic. Thus from the RDBMS perspective, representing dates in scalar form provides numerous benefits for users, both in terms of available operations and query performance.

Tableau models the first two of these types as "Date" and "Date \& Time"; the third (pure time) is folded into Date \& Time by appending it to a fixed date of \texttt{1899-12-30}. From the analytic perspective, date types are dimensional (\textit{i.e.} independent variables) and can be used as either categorical (simply ordered) or quantitative (ordered with a distance metric) fields.

Categorical dates have a natural hierarchy associated with them generated by calendar binning. The visualization in Figure I1 shows an example of a bar chart employing binned categorical dates in a year/quarter hierarchy.

\subsubsection{Parsing}
One of the oldest data preparation problems observed in Tableau is the parsing of date scalars so that users can perform these kinds of analyses. Training materials from the early days of the company included examples of how to convert columns of integers in the form \texttt{yyyyMMdd} to date scalars. The documented solution at the time was to convert the integer to a string and perform some locale-dependent string operations before casting the string back to a date. Unfortunately, this approach had a number of problems:
\begin{itemize}
\item String operations are notoriously slow compared to scalar operations (typically 10-100x slower)
\item Default parsing of date formats is locale-dependent, and may not work when the workbook is shared across an international organization (e.g. between the US and European offices)
\item The expression code was hard to understand and maintain because it used a verbose, general-purpose string-handling syntax instead of a domain language.
\end{itemize}

@@

Our first solution to this problem was to add a new function to the Tableau calculation language called \texttt{DATEPARSE}. This function would take a string column and convert it to a datetime using a special purpose domain language for describing dates. Such functions exist in a number of databases (e.g. Oracle, Postgres and MySQL) and adding it to the TDE was straightforward, so it was a natural addition to the function library.

We chose the date parsing syntax defined by the International Components for Unicode (or ICU) project \cite{icu} for the common domain language because it mirrored what was available in the Tableau code base, and translating this syntax into the formats used by various database vendors (e.g. MySQL, Oracle, Postgres) was relatively painless. There are a few patches of non-overlapping functionality, but they tend to be obscure and we provide warnings when functionality is missing.

\subsubsection{	Usability}
After shipping this new function to our user base, we investigated how it was being applied in the field. We examined a few months of workbooks that had been uploaded to Public after the release to see if and how it was being used. We found that although there were a number of new calculations using \texttt{DATEPARSE}, the error rate for the domain language syntax was about 15\%. \texttt{DATEPARSE} was capable of solving the problem, and some users were able to discover it, but the syntax did not appear to be easy to use reliably.

\subsection{Automated Format Extraction}
One solution might have been to design a graphical environment that enabled users to construct valid patterns. This would have involved a substantial development investment with no guarantee that the result would be correct if the user misunderstood the environment. Instead, we have developed two algorithms for automating the derivation of the format string, each of which uses a different machine learning technique. Both algorithms result in over 90\% parsing accuracy -- and 100\% syntactic correctness because they are machine generated.

% vidya: this probably should be its own section
% hawkfish: Not sure it is big enough, but I can see the logic.

\subsection{Related Work}
Data preparation has been a known analytic bottleneck since at least the description of the Potter's Wheel system \cite{PottersWheel}. Since then, several other interactive data preparation systems have been proposed \cite{Wrangler,Refine}. While effective, these systems all make assumptions about possible date formats (e.g. the domain system in \cite{PottersWheel}), which we suggest are too restrictive for real world data.

The Minimum Descriptive Length technique was first proposed in \cite{mdl}.  We have built on the version described in \cite{PottersWheel}.

Related work on parsing languages, outlier detection.

### 2015-07-06 mdl.tex
Word => TeX formatting
Authors, fixed-width fonts, citations, dumb quotes.

The first algorithm is a Minimum Descriptive Length [7] approach derived from the domain system presented in [4]. We describe a number of extensions to their structure extraction system to support more complex redundancy, non-English locales, improved performance and date-specific pruning.

\subsection{Domains}
[4] presents an algorithm for deriving a common structure for a set of strings by breaking each string down into a sequence of domains. These domains are described by an interface that includes:
\begin{itemize}
\item A required inclusion function to test for membership in the domain (match)
\item An optional function to compute the number of values in the domain with a given length (cardinality)
\item An optional function to update statistics for the domain based on a given value (updateStatistics)
\item An optional function to prevent consideration of a domain that is redundant (isRedundantAfter).
\end{itemize}

@@

\subsection{Redundancy Extensions}
A difficulty in using this kind of structure extraction is that the algorithm for enumerating structures is exponential in the number of domains. This is especially true in the date format problem because there are identical domains (e.g. months and meridian hours), nearly identical domains (e.g. days and hours) and there are often no field delimiters (e.g. 2012Mar06134427). To handle this, we have extended the existing pruning API with two other sets of domain identifiers:
\begin{itemize}
\item A set of \textit{prunable} identifiers, which are not allowed to precede the domain. For example, once we have a month field, no other month fields should be generated. Each month domain therefore lists all the month domains in its prunable set.
\item A set of \textit{context} identifiers, one of which must have been previously generated before the domain is considered. For example, a meridian domain can only be generated once an hour field has been found, but there may be other intervening fields.
\end{itemize}

@@

\subsection{Global Pruning}
The pruning rules used for the structure extraction reduce the search space dramatically, but they are also contextual and can only look backwards. The domains also contain a fair amount of ambiguity that requires the application of domain knowledge. We therefore found it necessary to add some post-generation global pruning rules:
\begin{itemize}
\item The set of date parts cannot contain place value gaps (e.g. structures that have year and day without month are removed.)
\item Similarly, the set of time parts cannot contain place value gaps and must also be in place value order (times are never written in orders such as mhs.)
\item The existence of time parts cannot make dates incomplete (e.g. patterns like year-day-hour are removed.)
\item Two digit years require special handling. In particular, they cannot appear adjacent to a two-digit field if the structure contains punctuation. (This can come up in some small early 21st century year domains where a two-digit year can masquerade as almost any numeric field.)
\end{itemize}

@@

\subsection{Ranking}
MDL structure analysis naturally produces a ranked list of format candidates, but we have found that a number of other properties of the formats should be preferred over simple compactness:
\begin{itemize}
\item Since we have a set of samples, we can apply the candidate format to the strings to see how well it performs. Formats with fewer parse errors are preferred.
\item Date parts can be considered a place-value system, so we prefer “more significant” components (e.g. month-day-year over hour-minute-second).
\item If two formats from different locales give the same results, prefer the original column locale. The sample set may have missed an example where this could be important.
\item If the format has an ambiguous date order (e.g. all days are less than 12), then prefer the default date order of the locale. Again, the sample set may have missed a counterexample, so this is the best option.
\item Once these semantic preferences have been considered, we then prefer the more compact (MDL) representation.
The output of the algorithm is now an ordered list of formats and associated locales. These can then be used to drive a user interface that allows the user to choose between the possibilities or the top-ranking format can simply be used automatically.
\end{itemize}

>>>>

The first algorithm is a Minimum Descriptive Length \cite{mdl} approach derived from the domain system presented in \cite{PottersWheel}. We describe a number of extensions to their structure extraction system to support more complex redundancy, non-English locales, improved performance and date-specific pruning.

\subsection{Domains}
\cite{PottersWheel} presents an algorithm for deriving a common structure for a set of strings by breaking each string down into a sequence of domains. These domains are described by an interface that includes:
\begin{itemize}
\item A required inclusion function to test for membership in the domain (\texttt{match})
\item An optional function to compute the number of values in the domain with a given length (\texttt{cardinality})
\item An optional function to update statistics for the domain based on a given value (\texttt{updateStatistics})
\item An optional function to prevent consideration of a domain that is redundant (\texttt{isRedundantAfter}).
\end{itemize}

@@

\subsection{Redundancy Extensions}
A difficulty in using this kind of structure extraction is that the algorithm for enumerating structures is exponential in the number of domains. This is especially true in the date format problem because there are identical domains (e.g. months and meridian hours), nearly identical domains (e.g. days and hours) and there are often no field delimiters (e.g. \texttt{2012Mar06134427}). To handle this, we have extended the existing pruning API with two other sets of domain identifiers:
\begin{itemize}
\item A set of \textit{prunable} identifiers, which are not allowed to precede the domain. For example, once we have a month field, no other month fields should be generated. Each month domain therefore lists all the month domains in its prunable set.
\item A set of \textit{context} identifiers, one of which must have been previously generated before the domain is considered. For example, a meridian domain can only be generated once an hour field has been found, but there may be other intervening fields.
\end{itemize}

@@

\subsection{Global Pruning}
The pruning rules used for the structure extraction reduce the search space dramatically, but they are also contextual and can only look backwards. The domains also contain a fair amount of ambiguity that requires the application of domain knowledge. We therefore found it necessary to add some post-generation global pruning rules:
\begin{itemize}
\item The set of date parts cannot contain place value gaps (e.g. structures that have year and day without month are removed.)
\item Similarly, the set of time parts cannot contain place value gaps and must also be in place value order (times are never written in orders such as \texttt{mhs}.)
\item The existence of time parts cannot make dates incomplete (e.g. patterns like year-day-hour are removed.)
\item Two digit years require special handling. In particular, they cannot appear adjacent to a two-digit field if the structure contains punctuation. (This can come up in some small early 21st century year domains where a two-digit year can masquerade as almost any numeric field.)
\end{itemize}

@@

\subsection{Ranking}
MDL structure analysis naturally produces a ranked list of format candidates, but we have found that a number of other properties of the formats should be preferred over simple compactness:
\begin{itemize}
\item Since we have a set of samples, we can apply the candidate format to the strings to see how well it performs. Formats with fewer parse errors are preferred.
\item Date parts can be considered a place-value system, so we prefer "more significant" components (e.g. month-day-year over hour-minute-second).
\item If two formats from different locales give the same results, prefer the original column locale. The sample set may have missed an example where this could be important.
\item If the format has an ambiguous date order (e.g. all days are less than 12), then prefer the default date order of the locale. Again, the sample set may have missed a counterexample, so this is the best option.
\item Once these semantic preferences have been considered, we then prefer the more compact (MDL) representation.
The output of the algorithm is now an ordered list of formats and associated locales. These can then be used to drive a user interface that allows the user to choose between the possibilities or the top-ranking format can simply be used automatically.
\end{itemize}

### 2015-07-06 paper.tex
Word => TeX formatting
Authors, fixed-width fonts, citations, dumb quotes.

@@

\title{Does Anybody Really Know What Time It Is?}
\subtitle{Automating the Extraction of Date Scalars}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

\numberofauthors{3} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
\alignauthor
Richard Wesley \\
%       \affaddr{Institute for Clarity in Documentation}\\
%       \affaddr{1932 Wallamaloo Lane}\\
%       \affaddr{Wallamaloo, New Zealand}\\
%       \email{trovato@corporation.com}
% 2nd. author
\alignauthor
Vidya Setlur\\
%       \affaddr{Institute for Clarity in Documentation}\\
%       \affaddr{P.O. Box 1212}\\
%       \affaddr{Dublin, Ohio 43017-6221}\\
%       \email{webmaster@marysville-ohio.com}
% 3rd. author
\alignauthor
Dan Cory\\
%       \affaddr{The Th{\o}rv{\"a}ld Group}\\
%       \affaddr{1 Th{\o}rv{\"a}ld Circle}\\
%       \affaddr{Hekla, Iceland}\\
%       \email{larst@affiliation.org}
}
% There's nothing stopping you putting the seventh, eighth, etc.
% author on the opening page (as the 'third row') but we ask,
% for aesthetic reasons that you place these 'additional authors'
% in the \additional authors block, viz.
\additionalauthors{Additional authors: John Smith (The Th{\o}rv{\"a}ld Group,
email: {\texttt{jsmith@affiliation.org}}) and Julius P.~Kumquat
(The Kumquat Consortium, email: {\texttt{jpkumquat@consortium.net}}).}
\date{30 July 1999}
% Just remember to make sure that the TOTAL number of authors
% is the number that will appear on the first page PLUS the
% number that will appear in the \additionalauthors section.

@@

One of the most common parsing tasks is extracting date and time data from string representations. Several databases include date parsing ``mini-languages" to cover the wide range of possible formats. Tableau has recently provided a \\DATEPARSE function with a single syntax that is translated into the nearest equivalent in the underlying database. Subsequent analysis of customer usage of this function shows that the parsing language syntax was difficult for users to master.

@@

%ACKNOWLEDGMENTS are optional
\section{Acknowledgments}
We would like to thank to Pat Hanrahan for encouraging this work; Roger Hau for providing the resources to productize it; and Jason King, whose harrowing live demonstration of manual DATEPARSE in front of several thousand people inspired this project.

>>>>

%
\def\sharedaffiliation{%
\end{tabular}
\begin{tabular}{c}}
%

@@

\conferenceinfo{SIGMOD}{'16 SanFrancisco, CA USA}

\title{Does Anybody Really Know What Time It Is?\\
Automating the Extraction of Date Scalars}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

\numberofauthors{3} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
\alignauthor
Richard Wesley \\
%       \affaddr{Tableau Software}\\
%       \affaddr{837 North 34th Street}\\
%       \affaddr{Seattle, WA 98103}\\
%      \email{hawkfish@tableau.com}
% 2nd. author
\alignauthor
Vidya Setlur\\
%       \affaddr{Tableau Software}\\
 %      \affaddr{837 North 34th Street}\\
%       \affaddr{Seattle, WA 98103}\\
 %      \email{vsetlur@tableau.com}
% 3rd. author
\alignauthor
Dan Cory\\
%       \affaddr{Tableau Software}\\
%       \affaddr{837 North 34th Street}\\
%       \affaddr{Seattle, WA 98103}\\
%       \email{dcory@tableau.com}
%       \email{\{hawkfish,vsetlur,dcory\}@tableau.com}
\sharedaffiliation
       \affaddr{Tableau Software}\\
       \affaddr{837 North 34th Street}\\
       \affaddr{Seattle, WA 98103}\\
       \email{\{hawkfish,vsetlur,dcory\}@tableau.com}
}
% Just remember to make sure that the TOTAL number of authors
% is the number that will appear on the first page PLUS the
% number that will appear in the \additionalauthors section.

@@

One of the most common parsing tasks is extracting date and time data from string representations. Several databases include date parsing ``mini-languages" to cover the wide range of possible formats. Tableau has recently provided a function called \texttt{DATEPARSE}  with a single syntax that is translated into the nearest equivalent in the underlying database. Subsequent analysis of customer usage of this function shows that the parsing language syntax was difficult for users to master.

@@

%ACKNOWLEDGMENTS are optional
\section{Acknowledgments}
We would like to thank to Pat Hanrahan for encouraging this work; Roger Hau for providing the resources to productize it; and Jason King, whose harrowing live demonstration of manual \texttt{DATEPARSE} in front of several thousand people inspired this project.

### 2015-07-06 parameters.tex
Word => TeX formatting
Authors, fixed-width fonts, citations, dumb quotes.

\subsection{Input Data}
The training data and test data are taken from Public data sets. We selected a large number of string columns whose names suggested that they might contain temporal data. We included words from multiple languages besides English and extracted the column’s locale (actually the collation) along with the data.

\subsubsection{NULL Filtering}
Each column was then reduced to a maximum sample set of 32. We excluded domain values that appeared to be representations of NULL:
\begin{itemize}
\item Values containing the substring NULL
\item Values containing no digits and at most one non-whitespace character (e.g. " / / ")
\item Values on a list of empirically determined common NULL values (e.g. 0000-00-00, NaN)
\end{itemize}

Columns that had no remaining valid samples were discarded.

\subsubsection{Sampling}
The remaining samples were hashed and sorted on the hash value, with the top 32 values being retained as the column’s sample. We can increase the sample size if needed, but for dates, it appears to be an adequate number. In any case, the median number of non-null rows per domain in our test data is 50, so increasing the sample size would have little benefit.

\subsubsection{Numeric Timestamps}
After NULL filtering, there remained one common class of date representation that was unsuited for parsing via our date format syntax, namely numeric timestamps. This included Unix epoch timestamps (expressed as second, millisecond or microsecond counts from 1970-01-01) and Microsoft Excel timestamps (expressed fractional days since 1900-01-01). A simple test to check that the column was numeric and inside a specific range of values representing recent dates allowed us to tag these columns to avoid analyzing them further (and incidentally to identify them for generating a simple date extraction calculation for the user.)

@@

The syntax is documented at the ICU web site [8]. While fairly complete, it has a few limitations that we ran into when evaluating the algorithms:
\begin{itemize}
\item No support for 4 letter year abbreviations (e.g. Sept.)
\item No support for ordinal days (e.g. July 4th)
\item No support for quarter postfix notation (e.g. 2Q)
\item No support for variant meridian markers (e.g. a.m.)
\end{itemize}

@@

One other quirk of the ICU syntax may be a contributing factor to the user confusion around writing correct ICU date formats. The use of lexicographical case in the meta-symbols of the format language can be confusing (e.g. y is used for years, but M is used for months while m is used for minutes.) Another advantage of an automated algorithm is that it hides such problems from most users, significantly improving the usability of the function.

@@

ICU’s date parsing APIs allow the specification of default values for parts when a format does not contain them. In our implementation, all time fields are set to 0 (midnight) and the date fields are set based on whether the format contains any date part specifications. When date parts are present, we use 2000-01-01 as the set of default date parts as it is the start of a leap year. When dealing with pure time formats, we use 1899-12-30, (which is the date used internally by Tableau to signal pure time values.)

>>>>

\subsection{Input Data}
The training data and test data are taken from Public data sets. We selected a large number of string columns whose names suggested that they might contain temporal data. We included words from multiple languages besides English and extracted the column's locale (actually the collation) along with the data.

\subsubsection{NULL Filtering}
Each column was then reduced to a maximum sample set of 32. We excluded domain values that appeared to be representations of \texttt{NULL}:
\begin{itemize}
\item Values containing the substring \texttt{NULL}
\item Values containing no digits and at most one non-whitespace character (e.g. \texttt{" / / "})
\item Values on a list of empirically determined common \texttt{NULL} values (e.g. \texttt{0000-00-00}, \texttt{NaN})
\end{itemize}
Columns that had no remaining valid samples were discarded.

\subsubsection{Sampling}
The remaining samples were hashed and sorted on the hash value, with the top 32 values being retained as the column's sample. We can increase the sample size if needed, but for dates, it appears to be an adequate number. In any case, the median number of non-null rows per domain in our test data is 50, so increasing the sample size would have little benefit.

\subsubsection{Numeric Timestamps}
After NULL filtering, there remained one common class of date representation that was unsuited for parsing via our date format syntax, namely numeric timestamps. This included Unix epoch timestamps (expressed as second, millisecond or microsecond counts from \texttt{1970-01-01}) and Microsoft Excel timestamps (expressed fractional days since \texttt{1900-01-01}). A simple test to check that the column was numeric and inside a specific range of values representing recent dates allowed us to tag these columns to avoid analyzing them further (and incidentally to identify them for generating a simple date extraction calculation for the user.)

@@

The syntax is documented at the ICU web site \cite{icu}. While fairly complete, it has a few limitations that we ran into when evaluating the algorithms:
\begin{itemize}
\item No support for 4 letter year abbreviations (e.g. \texttt{Sept.})
\item No support for ordinal days (e.g. \texttt{July 4th})
\item No support for quarter postfix notation (e.g. \texttt{2Q})
\item No support for variant meridian markers (e.g. \texttt{a.m.})
\end{itemize}

@@

One other quirk of the ICU syntax may be a contributing factor to the user confusion around writing correct ICU date formats. The use of lexicographical case in the meta-symbols of the format language can be confusing (e.g. \texttt{y} is used for years, but \texttt{M} is used for months while \texttt{m} is used for minutes.) Another advantage of an automated algorithm is that it hides such problems from most users, significantly improving the usability of the function.

@@

ICU's date parsing APIs allow the specification of default values for parts when a format does not contain them. In our implementation, all time fields are set to \texttt{0} (midnight) and the date fields are set based on whether the format contains any date part specifications. When date parts are present, we use \texttt{2000-01-01} as the set of default date parts as it is the start of a leap year. When dealing with pure time formats, we use \texttt{1899-12-30}, (which is the date used internally by Tableau to signal pure time values.)
